Modelling BCG vaccination in the UK

### 2019-08-07 rmds/08-model-development.Rmd
acronyms updated
upgraded model fitting result figs

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes via model simulation.

@@

## TB disease

@@

1. **Pulmonary/Extra-Pulmonary TB** - active TB disease can be defined as any symptomatic TB infection but it may present with a range of diverse individual states. Commonly, TB cases are stratified into pulmonary and extra-pulmonary TB cases, with pulmonary cases being individuals who present with TB present in the lungs, and extra-pulmonary TB cases being cases that present with TB symptoms that do not involve the lungs (Chapter \@ref(background)). Often pulmonary cases also present with extra-pulmonary symptoms. It is thought that pulmonary TB cases make up for the vast majority of TB transmission,[@Mathema2018; @Sepkowitz1996] as TB is primarily spread by aerosol transmission but that extra-pulmonary cases have worse outcomes. The proportion of pulmonary to extra-pulmonary cases has increased over time from 26.2% (1944/7410) in 1982 to 45.8% (2634/5748) in 2016. This may be attributed to the age distribution of TB cases changing, as different age-groups are more likely to progress to pulmonary vs extra-pulmonary TB.[@Lefebvre2017] It may also be related to the increase of non-UK born cases, as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016) [@PHE2017]. The model presented here includes both pulmonary and extra-pulmonary cases, with only pulmonary cases contributing to onwards transmission. Extra-pulmonary cases are included so that the full impacts of any intervention can be correctly estimated.

@@

1. **Non-UK born TB Cases** - TB incidence in England is highly heterogeneous with over 70% of cases occurring in the non-UK born population.[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016).[@PHE2017] Exposure to England's BCG vaccination policy is difficult to assess for the non-UK born as is the degree of transmission occurring in the UK as opposed to cases being imported from abroad, or acquired from trips to cases countries of origin. For this reason the model presented here does not explicitly include non-UK born cases. Instead it imports non-UK born cases into the force of infection,[@Anderson1991; @Keeling2007] with the addition of a mixing parameter that controls the degree of contact between non-UK born cases and those born in the UK.

@@

1. **Protection from active disease** - the BCG vaccine has been shown to primarily protect against the progression from latent to active TB disease (Chapter \@ref(background)). It has been shown to be highly protective in children,[@Rodrigues1993; @Colditz1994; @Roy2014] but to have variable protection in adults ranging from 0-80%.[Mangtani2014; @Zwerling2011] This variation in protection is thought to be linked to the equator, with the vaccine becoming increasing effective at higher, and lower latitudes. In England an MRC trial in the 1950's found that the BCG vaccine was highly effective, there is little evidence to suggest that this has changed in the UK born population.[@Hart1972]

1. **Duration of protection** - BCG protection wanes with time, with the greatest protection shortly after vaccination. There is good evidence to suggest that the effectiveness of BCG vaccination lasts up to 15 years,[@Abubakar2013] and a recent study suggests that this protection may last later into adulthood in the UK born.[@Mangtani2017] However, the it's effectiveness reduces with time since vaccination.

@@

The dynamic model of TB implemented here may be considered as 3 nested models; these are a TB transmission model; a demographic processes model; and a BCG vaccination model. For an overview of the model structure see the flow diagram (Figure \@ref(fig:model-flow-diag)) and for full details see the model equations (Section \@ref(model-equations)). Model parameters are discussed in detail in Section \@ref(model-parameters).

@@

The model is stratified into 11 age groups with 5 year age groups from 0 to 49, a single age group from 50-69, and a single age group from 70 to 89. Older adults were grouped into larger age groups as they are thought to be responsible for a small amount of TB transmission and because fine scale BCG mechanisms do not need to be modelled in these age groups. Adults aged 90+ were not modeled due to large amounts of uncertainty in the demographic data and because cases in this population represent a small fraction of total TB cases (see Chapter \@ref(data)). The number of births in a given year is incorporated as a time varying parameter. The natural mortality rate is also allowed to vary with time and is stratified by age. Immigration and emigration were not included in the demographic model as reliable age stratified data was unavailable and it is unlikely that either immigration or emigration of the UK born population is a significant driver of overall population size, or structure.

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the age-specific effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the age-specific protection inferred due to vaccination from initial infection. In the unvaccinated population $\alpha = 0$ and $\chi = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

@@

```{r disease-model, eval = FALSE}
disease_params_table %>% 
  select(Parameter, Description, Distribution, Units, Method, Type) %>% 
kable(
  caption.short = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type.", 
  caption = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type (i.e data derived, literature, assumption). All data based parameters are included. P = pulmonary TB, E = extra-pulmonary TB, v = vaccinated, i = age at vaccination, $\\mathcal{U}$ = Uniform, $\\mathcal{N}$ = Normal", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header", "hold_position")) %>% 
  column_spec(c(5), width = "6cm") %>% 
  column_spec(c(2), width = "4cm") %>%
  column_spec(c(3), width = "6cm") %>%
  column_spec(c(1,4,6), width = "1.5cm") %>% 
  landscape()
```

@@

sources_table %>% 
  left_join(params_and_sources, by = "Source") %>% 
  select(Parameters, `Study Type`, Setting, Year, Description, Source) %>% 
  pander(caption = "(\\#tab:sources-tab) Sources used to parameterise the disease and demographic models. Parameters that use the source are given, as well as the study type, setting, year/years studied and a description of the study/data source.", split.cells = c(8, 15, 10, 8, 8, 40, 6), split.table = Inf)
```

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-69) older adults (70-89). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases. Vertical lines indicate the 2.5%, 25%, 50%, 75%, and 97.5% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups;children (0-14), adults (15-69) older adults (70-89). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times. Vertical lines indicate the 2.5%, 25%, 50%, 75%, and 97.5% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

>>>>

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes, via model simulation.

@@

### TB disease

@@

1. **Pulmonary/Extra-Pulmonary TB** - active TB disease can be defined as any symptomatic TB infection but it may present with a range of diverse individual states. Commonly, TB cases are stratified into pulmonary and extra-pulmonary TB cases, with pulmonary cases being individuals who present with TB present in the lungs, and extra-pulmonary TB cases being cases that present with TB symptoms that do not involve the lungs (Chapter \@ref(background)). Often pulmonary cases also present with extra-pulmonary symptoms. It is thought that pulmonary TB cases make up for the vast majority of TB transmission,[@Mathema2018; @Sepkowitz1996] as TB is primarily spread by aerosol transmission, but that extra-pulmonary cases have worse outcomes. The proportion of pulmonary to extra-pulmonary cases has increased over time from 26.2% (1944/7410) in 1982 to 45.8% (2634/5748) in 2016. This may be attributed to the age distribution of TB cases changing, as different age-groups are more likely to progress to pulmonary vs extra-pulmonary TB.[@Lefebvre2017] It may also be related to the increase of non-UK born cases, as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] The model presented here includes both pulmonary and extra-pulmonary cases, with only pulmonary cases contributing to onwards transmission. Extra-pulmonary cases are included so that the full impacts of any intervention can be correctly estimated.

@@

1. **Non-UK born TB Cases** - TB incidence in England is highly heterogeneous with over 70% of cases occurring in the non-UK born population.[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75-79 years old (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016).[@PHE2017] Exposure to England's BCG vaccination policy is difficult to assess for the non-UK born as is the degree of transmission occurring in the UK as opposed to cases being imported from abroad, or acquired from trips to cases countries of origin. For this reason the model presented here does not explicitly include non-UK born cases. Instead it imports non-UK born cases into the force of infection with a mixing parameter that controls the degree of contact between non-UK born cases and those born in the UK.[@Anderson1991; @Keeling2007]

@@

1. **Protection from active disease** - the BCG vaccine has been shown to primarily protect against the progression from latent to active TB disease (Chapter \@ref(background)). It has been shown to be highly protective in children[@Rodrigues1993; @Colditz1994; @Roy2014] but to have variable protection in adults ranging from 0-80%.[Mangtani2014; @Zwerling2011] This variation in protection is thought to be linked to the equator with the vaccine becoming increasing effective at higher, and lower latitudes. In England, an MRC trial in the 1950's found that the BCG vaccine was highly effective.[@Hart1972] There is little evidence to suggest that this has changed in the UK born population.

1. **Duration of protection** - BCG protection wanes with time, with the greatest protection shortly after vaccination. There is good evidence to suggest that the effectiveness of BCG vaccination lasts up to 15 years,[@Abubakar2013] and a recent study suggests that this protection may last later into adulthood in the UK born.[@Mangtani2017]

@@

The dynamic model of TB implemented here may be considered as 3 nested model these are: a TB transmission model; a demographic processes model; and a BCG vaccination model. For an overview of the model structure see the flow diagram (Figure \@ref(fig:model-flow-diag)) and for full details see the model equations (Section \@ref(model-equations)). Model parameters are discussed in detail in Section \@ref(model-parameters).

@@

The model is stratified into 11 age groups with 5 year age groups from 0 to 49, a single age group from 50-69, and a single age group from 70 to 89. Older adults were grouped into larger age groups as they are thought to be responsible for a small amount of TB transmission and because fine scale BCG mechanisms do not need to be modelled in these age groups. Adults aged 90+ were not modeled due to large amounts of uncertainty in the demographic data and because cases in this population represent a small fraction of total TB cases (see Chapter \@ref(data)). The number of births in a given year is incorporated as a time varying parameter. The natural mortality rate is also allowed to vary with time and is stratified by age. Immigration and emigration were not included in the demographic model as reliable age stratified data were unavailable and it is unlikely that either immigration or emigration of the UK born population is a significant driver of overall population size, or structure.

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the protection inferred due to vaccination from initial infection. In the unvaccinated population $\alpha = 0$ and $\chi = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

@@

```{r disease-model}
disease_params_table %>% 
  select(Parameter, Description, Distribution, Units, Method, Type) %>% 
kable(
  caption.short = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type.", 
  caption = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type (i.e data derived, literature, assumption). All data based parameters are included. P = pulmonary TB, E = extra-pulmonary TB, v = vaccinated, i = age at vaccination, $\\mathcal{U}$ = Uniform, $\\mathcal{N}$ = Normal", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header", "hold_position")) %>% 
  column_spec(c(5), width = "6cm") %>% 
  column_spec(c(2), width = "4cm") %>%
  column_spec(c(3), width = "6cm") %>%
  column_spec(c(1,4,6), width = "1.5cm") %>% 
  landscape()
```

@@

sources_table %>% 
  left_join(params_and_sources, by = "Source") %>% 
  select(Parameters, `Study Type`, Setting, Year, Description, Source) %>% 
  pander(caption = "(\\#tab:sources-tab) Sources used to parameterise the disease and demographic models. Parameters that use the source are given, as well as the study type, setting, year/years studied and a description of the study/data source.", split.cells = c(4, 10, 10, 8, 8, 50, 6), split.table = Inf)
```

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-69) older adults (70-89). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases. Vertical lines indicate the 2.5\\%, 25\\%, 50\\%, 75\\%, and 97.5\\% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups;children (0-14), adults (15-69) older adults (70-89). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times. Vertical lines indicate the 2.5\\%, 25\\%, 50\\%, 75\\%, and 97.5\\% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

### 2019-08-07 rmds/09-model-fitting.Rmd
updated model fitting
upgraded model fitting result figs

Table \@ref(tab:09-scenarios-table)
```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
  str_to_sentence()) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

```{r 09-fig-sens, fig.scap = "Partial rank correlation coefficients for each parameter fitted too.", fig.cap = "Partial rank correlation coefficients for each parameter fitted too. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$, $c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$", out.width = "300px"}
knitr::include_graphics(file.path(plots_path, "incidence-sensitivity-1.png"))
```

@@

Whilst the results presented here are not encouraging it may be the case that with a greatly extended run-time a better fitting model may be found. The first step to testing this is to rerun the pipeline using several orders of magnitude more rejuvenation steps - as these have no RAM overhead. Even if the resulting model fits are still poor this may indicate areas for improvement - in either the model fitting pipeline or the model itself. Reducing model complexity may also help via decreasing the computational burden and so allowing more particles to be used. An alternative option is to re-code the model developed in the last chapter into a more generic form (i.e C) and to attempt to fit the model using other techniques that are less compute intensive and more robust - such as SMC-ABC. Another option would be to re-implement the model in the form required by another modelling package - such as `pomp`. However, this would mean running the risk of again getting stuck within a framework that is difficult to debug and may not be working as expected. A final option would be to reduce the complexity of the model and potentially fit to less complex data. This may be the only solution if the current model is not identifiable.

>>>>

\newpage

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
  str_to_sentence()) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

\newpage

```{r 09-fig-sens, fig.scap = "Partial rank correlation coefficients for each parameter fitted too.", fig.cap = "Partial rank correlation coefficients for each parameter fitted too. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$, $c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$", out.width = "250px"}
knitr::include_graphics(file.path(plots_path, "incidence-sensitivity-1.png"))
```

@@

Whilst the results presented here are not encouraging it may be the case that with a greatly extended run-time a better fitting model may be found. The first step to testing this is to rerun the pipeline using several orders of magnitude more rejuvenation steps - as these have no RAM overhead. Even if the resulting model fits are still poor this may indicate areas for improvement - in either the model fitting pipeline or the model itself. Reducing model complexity may also help via decreasing the computational burden and so allowing more particles to be used. Possible simplifications are to remove features, such as the treatment population, that do little to alter the dynamics but are present due to use case concerns or switch from a continous framework to a discrete one. An alternative option is to re-code the model developed in the last chapter into a more generic form (i.e C) and to attempt to fit the model using other techniques that are less compute intensive and more robust - such as SMC-ABC. Another option would be to re-implement the model in the form required by another modelling package - such as `pomp`. However, this would mean running the risk of again getting stuck within a framework that is difficult to debug and may not be working as expected. A final option would be to reduce the complexity of the model and potentially fit to less complex data. This may be the only solution if the current model is not identifiable.
